Measuring contrast processing in the visual system using SSVEP

### 2024-05-06 review.qmd
Minor edits

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006, @Muller1998m @MullerHillyard2000, @MorganPNAS1996, @Hillyard1998NNS].

Electrophysiological and psychophysical measurements of the effect of attention on both luminance and chromatic contrast have strongly implicated gain control as an underlying mechanism [@DiRusso1999; @DiRusso2001; @Lu1998]. These effects also appear to differ between chromatic and achromatic pathways [@DiRusso1999b] - perhaps as a result of the different levels of nonlinear gain control in the early pre-cortical magno-, parvo- and konio-cellular pathways [@Derrington1984b; @Kaplan1986; @Lee1990; @Solomon2005]. In the late 2000s a comprehensive theoretical model for attentional modulation was developed that framed it as a gain control computation [@Reynolds2009; @Boynton2009]. This framework has proven to be influential - explaining a wide range of phenomena from the earlier literature and demonstrating subtle interactions between the size of the attentional 'spotlight' and the stimulus configuration which rationalise many apparent contradictions in the literature. Direct measurements of attentional modulation of achromatic SSVEP signals are broadly consistent with this model [@Lauritzen2010, @Hou2016, @Martinovic2018] and confirm the relatively weaker role of spatial attention on responses driven by chromatic stimuli - particularly those that isolate the opponent S-(L+M) cone pathway [@Wang2011, @Highsmith2010]. The gain control model of attention can also be extended to SSVEP studies of feature-based attention which show that the modulatory effects can be targeted to the most informative neuronal populations [@Verghese2012].

The SSVEP can also be used to study changes in visual processing \textit{driven} by different behavioural states or overall arousal. For example, locomotion has been shown to alter neuronal excitability and spatial normalization in mice [@Niell2010; @Ayaz2013] - running mice have higher visual sensitivity and lower surround suppression compared to stationary mice. Although measuring EEG responses from locomoting humans is technically challenging, SSVEP studies (which are able to distinguish broadband noise from input signal effectively) have shown that walking also alters early visual processing although in a manner different to that observed in mice [@Benjamin2018; @Cao2019].

SSVEP measurements are typically used to measure time invariant responses due to sustained attention. However, recent work has shown that moderately high modulation frequencies (ca 10Hz) and short analysis windows (1.5s) can also be used to track the dynamic allocation of attention across a task [@Chota2024] or attention to moving targets [@Lissa2020]. The ability to track changes in attention over short time periods is also important if SSVEP is to be used for dynamic readouts - for example in a Brain Computer Interface.

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010],  the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is typically present, this modulation and the associated changes in gain control, will affect the entire pattern of self- and intermodulation terms allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Conclusions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants) and at frequencies high enough to be effectively invisible [REF - Some high frequency work... - Romanian group]. The SSVEP is also able 'fingerprint' modulators of the input frequency through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by

>>>>

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006; @Muller2000; @Morgan1996; @Muller1998].

Electrophysiological and psychophysical measurements of the effect of attention on both luminance and chromatic contrast have strongly implicated gain control as an underlying mechanism [@DiRusso1999; @DiRusso2001; @Lu1998]. These effects also appear to differ between chromatic and achromatic pathways [@DiRusso1999b] - perhaps as a result of the different levels of nonlinear gain control in the early pre-cortical magno-, parvo- and konio-cellular pathways [@Derrington1984b; @Kaplan1986; @Lee1990; @Solomon2005]. In the late 2000s a comprehensive theoretical model for attentional modulation was developed that framed it as a gain control computation [@Reynolds2009; @Boynton2009]. This framework has proven to be influential - explaining a wide range of phenomena from the earlier literature and demonstrating subtle interactions between the size of the attentional 'spotlight' and the stimulus configuration which rationalise many apparent contradictions in the literature. Direct measurements of attentional modulation of achromatic SSVEP signals are broadly consistent with this model [@Lauritzen2010; @Hou2016; @Martinovic2018] and confirm the relatively weaker role of spatial attention on responses driven by chromatic stimuli - particularly those that isolate the opponent S-(L+M) cone pathway [@Wang2011; @Highsmith2010]. The gain control model of attention can also be extended to SSVEP studies of feature-based attention which show that the modulatory effects can be targeted to the most informative neuronal populations [@Verghese2012].

The SSVEP can also be used to study changes in visual processing \textit{driven} by different behavioural states or overall arousal. For example, locomotion has been shown to alter neuronal excitability and spatial normalization in mice [@Niell2010; @Ayaz2013] - running mice have higher visual sensitivity and lower surround suppression compared to stationary mice. Although measuring EEG responses from locomoting humans is technically challenging, SSVEP studies (which are able to distinguish broadband noise from input signal effectively) have shown that walking also alters early visual processing, although in a manner different to that observed in mice [@Benjamin2018; @Cao2019].

SSVEP measurements are typically used to measure time-invariant responses due to sustained attention. However, recent work has shown that moderately high modulation frequencies (ca 10Hz) and short analysis windows (1.5s) can also be used to track the dynamic allocation of attention across a task [@Chota2024] or attention to moving targets [@Lissa2020]. The ability to track changes in attention over short time periods is also important if SSVEP is to be used for dynamic readouts - for example in a Brain Computer Interface.

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and precise frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010], the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is generally present, this modulation and the associated changes in gain control will affect the entire pattern of self- and intermodulation terms, allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties: for example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Conclusions

The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants) and at frequencies high enough to be effectively invisible [@Seijdel2023]. The SSVEP is also able 'fingerprint' modulators of the input frequency through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by
